---
title: "HISP NASS Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Appendix 1: Data from MySQL:

For each dataset, the query was saved into seperate .CSV files, uploaded to GitHub, and imported into MySQL. From MySQL they were prepped to be ready to import into R. This section with detail the data files (original .CSV's), the source, and the queries that were used to create our final datasets: *honeystate* and *honeycounty*. The queries past the import wizard can be found here, and the query has been saved in [GitHub](add link to MySQL query).

### Data Import:

Simply enough, each .CSV was imported into a schema (titled **HISP**) using the MySQL data import wizard.

#### National Agriculture Statistics Service:

The first set of data we gathered came from the [National Agriculture Statistics Service website](https://quickstats.nass.usda.gov/), which houses a large amount of data related to United States Agriculture. The website stores census and survey data on a number of subjects (the main categories are animals, products, crops, demographics, economics, and environment) across a number of geographic levels (national, state, county, and zip code) for a number of years. The data can be access by the website UI allowing the user to select each subject, geographic level, and year which can then be exported into a .CSV file, or the entire database can be downloaded as a text file after extracting the database .GZ. For our purposes, it was simple enough to query the database using the website rather than have to deal with the large and cumbersome dataset. 

If we were to import the data into R directly we'd use the following (the [SQL query can be found using this link](https://github.com/chrisgmartin/HoneyIShrunkTheBees/blob/master/HoneyIShrunkTheBees.sql) ):

```{r eval=FALSE}
#County Data
county1 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/CAHoneyCensus_County_2012.csv", header = TRUE) #Honey Census 2012 with no other years
county2 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAChemicalandFertilizerTotals_County_allyears.csv", header = TRUE)
county3 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USACropTotalsCensus_County_allyears.csv", header = TRUE)
county4 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAHoneyCensus_County_allyears.csv", header = TRUE)
county5 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAHorticultureCensus_County_allyears.csv", header = TRUE)

#State Data
state1 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAHoneySurvey_State_allyears.csv", header = TRUE) #State Survey (no county survey)
state2 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAChemicalandFertilizerTotals_State_allyears.csv", header = TRUE)
state3 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USACropTotalsCensus_State_allyears.csv", header = TRUE)
state4 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAHoneyCensus_State_allyears.csv", header = TRUE)
state5 <- read.csv("https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/USAHorticultureCensus_State_allyears.csv", header = TRUE)

#Final Data
honey_county <- read.csv('https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/honey_county.csv', header = TRUE)
honey_state <- read.csv('https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/honey_state.csv', header = TRUE)
FIPS <- read.csv('https://raw.githubusercontent.com/chrisgmartin/HoneyIShrunkTheBees/master/Data/FIPS.csv', header = TRUE)
```





However, since the data is now in MySQL we query it for the following table names:

- cahoneycensus_county_2012

- USAChemicalandFertilizerTotals_County_allyears

- USAChemicalandFertilizerTotals_State_allyears

- USACropTotalsCensus_County_allyears.csv

- USACropTotalsCensus_State_allyears.csv

- USAHoneyCensus_County_allyears.csv

- USAHoneyCensus_State_allyears.csv

- USAHoneySurvey_State_allyears.csv

- USAHorticultureCensus_County_allyears.csv

- USAHorticultureCensus_State_allyears.csv

And merge all of those datasets into two simple tables (views technically, so that we do not contaminate the original data) titled:

- honey_county: which lists all of the county level statistics

- honey_state: which lists all of the state level statistics

#### Data on Bee Colonies







#### Data on County IDs

Since the datasets containing county level information use the county name as the identifier, we need to add a county number to it for our maps.





